---
title: "Class_1"
author: "Or Duek"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Central Limit Theorm

Central Limit Theorm

The central limit theorem states that the distribution of sample averages approaches a normal distribution as the sample size gets larger

Now we will build a simulation showing the central limit theorem

We will take a random sample of size 10 from a uniform distribution

We will repeat this process 1000 times and calculate the sample average each time

We will then plot a histogram of the sample averages

We will then repeat the process with a sample size of 30

```{r}
# Set the seed
set.seed(123)

# generate normal distribution of 5000 samples
normal_dist <- rnorm(5000, mean = 0, sd = 1)
# plot
hist(normal_dist, breaks = 50, col = "lightblue", main = "Normal Distribution", xlab = "Value")

```

## Nice

-   Now lets play with it

```{r}
# Take a random sample of size 10 from the normal distribution
sample_10 <- sample(normal_dist, 10)
hist(sample_10, breaks = 50, col = "lightblue", main = "Sample of 10", xlab = "Value")
```

### Hmmm

Doesn't look so good, right?

```{r}
# now, leat take n=10 and generate average of them, repeat 1000 times
sample_means_10 <- replicate(1000, mean(sample(normal_dist, 10)))

# plot
hist(sample_means_10, breaks = 50, col = "lightblue", main = "Sample Means of 10", xlab = "Value")
```

Nicer, right?

ok, but this is normal distribution, what about uniform distribution?

```{r}

# Set the seed
set.seed(123)


# generate uniform distribution of 5000 samples
uniform_dist <- runif(5000, min = 0, max = 1)
hist(uniform_dist, breaks = 50, col = "lightblue", main = "Uniform Distribution", xlab = "Value")

# Take a random sample of size 10 from the uniform distribution
sample_10 <- sample(uniform_dist, 10)
hist(sample_10, breaks = 50, col = "lightblue", main = "Sample of 10", xlab = "Value")

```

### Let's take average and repeaet

```{r}
# now, leat take n=10 and generate average of them, repeat 1000 times
sample_means_10 <- replicate(1000, mean(sample(uniform_dist, 10)))

# plot
hist(sample_means_10, breaks = 50, col = "lightblue", main = "Sample Means of 10", xlab = "Value")
```

# BAM!!
