---
title: "bios_class_3"
author: "Or Duek"
date: "`r Sys.Date()`"
output: html_document
---

```{r}
# load libraries
library(ggplot2)
library(tidyverse)
library(gtsummary)
```

# Today we will discuss Correlation between two variables

- When we want to test the association between two continuous variables, we can use pearson's correlation. 
- Pearson has a number of assumptions:
  - The two variables are continuous
  - The two variables are normally distributed
  - The two variables have a linear relationship
  - Homoscedasticity


Let's simulate two correlating variables  

```{r}
var_a <- rnorm(20, mean = 10, sd = 2)
var_b <- var_a - rnorm(20, mean = 2, sd = 3)

# plot them 
plot(var_a, var_b)

```
  
  
## Lets test the correlation

```{r}
cor.test(var_a, var_b)
```

  
# A-parametric tests
Usually use the order

- Mann-whitney U test

```{r}
# compare age of visitors of two different doctors:
doc_a <- c(8.2,9.3,9.4,9.5,12.6,12.7,12.8)
doc_b <- c(9.1,9.2,12.5,13.8,17.9,17.9,19.3)

# plot them
plot(doc_a, doc_b)

# do mann-whitney U test (only Wilcoxon test is available in R, set Paired to False)
wilcox.test(doc_a, doc_b, paired = FALSE) # Mann-Whitney U test
```


- Wilcoxon signed rank test
When two samples are dependent (before/after) 
We do Wilcoxon signed rank test

```{r}
# compare sutents rating in two different professors
prof_a <- c(4,2,1,5,5,4,5,2,5,5)
prof_b <- c(2,1,2,4,5,5,4,2,2,3)

wilcox.test(prof_a, prof_b, paired = TRUE, correct = F) # Wilcoxon signed rank test
```

- Kruskal-Wallis test
When we have more than two groups, we can use Kruskal-Wallis test

```{r}
# simulate data
set.seed(123)
group_a <- rnorm(20, mean = 10, sd = 2)
group_b <- rnorm(20, mean = 12, sd = 2)
group_c <- rnorm(20, mean = 14, sd = 2)

# plot
boxplot(group_a, group_b, group_c)

# nicer plot with ggplot
df <- data.frame(
  value = c(group_a, group_b, group_c),
  group = rep(c("A", "B", "C"), each = 20)
)
ggplot(df, aes(x = group, y = value, color=group)) +
  geom_boxplot() + geom_jitter(width = 0.2, alpha = 0.5) +
  labs(title = "Kruskal-Wallis Test", x = "Group", y = "Value") +
  theme_minimal()
# run test
kruskal.test(list(group_a, group_b, group_c))

```
- Friedman test
When we have more than two groups and the samples are dependent (before/after)

```{r}
# simulate data
set.seed(3)
time_a <- rnorm(20, mean = 10, sd = 5)
time_b <- time_a + rnorm(20, mean = 2, sd = 2)
time_c <- time_b + rnorm(20, mean = 2, sd = 4)

# Create a subject/blocking factor
subject <- factor(rep(1:20, 3))

# Create dataframe with subject ID
df <- data.frame(
  subject = subject,
  value = c(time_a, time_b, time_c),
  time = factor(rep(c("A", "B", "C"), each = 20))
)

# Plot
library(ggplot2)
ggplot(df, aes(x = time, y = value, color = time)) +
  geom_boxplot() + geom_jitter(width = 0.2, alpha = 0.5) +
  labs(title = "Friedman Test", x = "Time", y = "Value") +
  theme_minimal()

# Friedman test with subject as a blocking variable
friedman.test(value ~ time | subject, data = df)

```

## Spearman's correlation
- Spearman's correlation is a non-parametric test that can be used to test the association between two continuous variables.

- When no normal distribution is assumed
- When small samples

```{r}
# simulate data
a <- rnorm(7, mean = 10, sd = 2)
b <- a + rnorm(7, mean = 2, sd = 3)

# plot
plot(a, b)

# run spearman's correlation
cor.test(a, b, method = "spearman")
```

