---
title: "bios_class_2"
author: "Or Duek"
date: "`r Sys.Date()`"
output: html_document
---

```{r}
# load libraries
library(ggplot2)
library(tidyverse)
library(gtsummary)
```

# Today we will discuss T tests and ANOVA

-   t test is meant to compare two groups (dependent or independent)

## Independent t test

-   assume that the groups are not dependent
-   assumes that the observed (i.e. dependent) variable is normally distributed
-   assumes that the groups have equal variances (i.e., homoschedasticity)

```{r}
# Independent t test
# lets start by simulate data of two groups of people (men/women) and their weight
set.seed(123)
# Define the sample size
n <- 100

# Simulate gender
gender <- sample(c("Men", "Women"), n, replace = TRUE)

# Simulate weight
weight <- ifelse(gender == "Men",
                 rnorm(n, mean = 80, sd = 5),   # Mean weight for men: 80kg
                 rnorm(n, mean = 70, sd = 5))   # Mean weight for women: 70kg

# Create data frame
data <- data.frame(Gender = gender, Weight = weight)

```

Plot it (Exploratory data analysis)

Table1

```{r}
tbl_summary(data,
            by= Gender)
```

Plot

```{r}
ggplot(data, aes(x=Gender, y= weight, color= Gender)) +
  geom_boxplot() +
  geom_jitter(width = 0.2, height = 0, alpha = 0.5) +
  theme_minimal()
```

Now lets run the t test

```{r}
# before we run t test, we need to test assumptions
# test that observed variable is normally distributed
shapiro.test(data$Weight)
# show qqplot
qqnorm(data$Weight)

```

```{r}
t.test(Weight ~ Gender, data = data, var.equal = T)
```

-   lest see if variance is equal

```{r}
library(car)
leveneTest(Weight ~ Gender, data=data)
```

## Paired t test

```{r}
# Paired t test
# lets simulate two measures of depressive symptoms before and after treatment
set.seed(124)
# Define the sample size
n <- 100
# lets say treatment had effect of average 5 points, with sd of 2
before <- rnorm(n, mean = 20, sd = 5)
after <- before + rnorm(n, mean = -5, sd = 2)

# Create data frame
data_paired <- data.frame(Before = before, After = after)
```

```{r}
hist(after)
```


```{r}
# show histograms of before (red) and after (blue)
df_long <- pivot_longer(data_paired, cols = c(Before, After), names_to = "Condition", values_to = "Value")

# Plotting histograms
ggplot(df_long, aes(x = Value, fill = Condition)) +
  geom_histogram(alpha = 0.5, position = "identity", bins = 15) +
  scale_fill_manual(values = c("Before" = "red", "After" = "blue")) +
  labs(title = "Histogram: Before vs After", x = "Value", y = "Frequency") +
  theme_minimal()
```

Cool, now lets t test it

```{r}
# paired t-test
paired_test <- t.test(data_paired$Before, data_paired$After, paired = T)
paired_test
```

```{r}
# Load ggplot2
library(ggplot2)

# Define your data
mean_value <- paired_test$estimate
ci_lower <- paired_test$conf.int[1]
ci_upper <- paired_test$conf.int[2]

# Create a data frame for plotting
df_ci <- data.frame(
  Measure = "Paired t-test",
  Mean = mean_value,
  LowerCI = ci_lower,
  UpperCI = ci_upper
)

# Plotting
ggplot(df_ci, aes(x = Measure, y = Mean)) +
  geom_point(size = 4, color = "darkblue") +
  geom_errorbar(aes(ymin = LowerCI, ymax = UpperCI), width = 0.1, color = "darkblue") +
  coord_flip(ylim=c(-1,7)) +
  labs(title = "Mean with 95% Confidence Interval",
       x = "",
       y = "Value") +
  theme_minimal(base_size = 14) +
  theme(panel.grid.major.y = element_blank(),
        axis.text.y = element_text(face = "bold"))

```

# Now we do Chi square test

```{r}
set.seed(42)  # reproducibility

# Define sample size per medication group
n <- 100

# Generate medication groups
medication <- rep(c("Med_A", "Med_B", "Med_C"), each = n)

# Simulate recovery (Med_B has higher effectiveness)
recovery <- c(
  rbinom(n, 1, 0.31),  # Med_A: 31% recovery rate
  rbinom(n, 1, 0.70),  # Med_B: 70% recovery rate (helps recovery)
  rbinom(n, 1, 0.35)  # Med_C: 35% recovery rate
)

# Create data frame
df <- data.frame(
  Medication = medication,
  Recovered = factor(recovery, labels = c("No", "Yes"))
)

# View data summary
table(df$Medication, df$Recovered)

```

Now we run the Chi square test

```{r}
chisq.test(table(df$Medication, df$Recovered))
```

```{r}
tbl <- table(df$Medication, df$Recovered)

# Chi-square test
chi_result <- chisq.test(tbl)

# Extract actual chi-square value and degrees of freedom
chi_value <- chi_result$statistic
df_chi <- chi_result$parameter

# Critical value (alpha = 0.05 significance level)
alpha <- 0.05
critical_value <- qchisq(1 - alpha, df_chi)

# Create data frame for plotting
x_vals <- seq(0, chi_value + 10, length.out = 500)
chi_density <- dchisq(x_vals, df = df_chi)

plot_df <- data.frame(x = x_vals, y = chi_density)

# Plot chi-square distribution
ggplot(plot_df, aes(x, y)) +
  geom_line(color = "darkblue", size = 1) +
  
  # Shade critical region
  geom_area(data = subset(plot_df, x > critical_value), fill = "red", alpha = 0.3) +
  
  # Add critical value line
  geom_vline(xintercept = critical_value, color = "red", linetype = "dashed", linewidth = 1) +
  annotate("text", x = critical_value, y = max(chi_density)*0.9, 
           label = paste("Critical Value\n", round(critical_value,2)), 
           color = "red", hjust = -0.1) +
  
  # Add actual chi-square statistic line
  geom_vline(xintercept = chi_value, color = "black", linetype = "solid", linewidth = 1) +
  annotate("text", x = chi_value, y = max(chi_density)*0.7, 
           label = paste("Actual Value\n", round(chi_value,2)), 
           color = "black", hjust = -0.1) +
  
  labs(title = "Chi-square Distribution with Critical and Actual Values",
       x = expression(chi^2 ~ "Statistic"),
       y = "Density") +
  theme_minimal(base_size = 14)
```

## Fisher's exact test

```{r}
# lets do Fisher's exact test
fisher.test(tbl)
```

## McNemar's test

```{r}
# lets simulate data for McNemar's test
df_mc <- matrix(c(29,13,39,19), nrow=2,
                dimnames = list("After Video" = c("Support", "No Support"),
                                "Before Video" = c("Support", "No Support")))

mcnemar.test(df_mc)
```

# ANOVA

-   What should we do when we have more than two groups to compare?

-   Analysis Of Variance

-   We expect the variance between the groups (MSG/MSB) to be larger than variance within the groups (MSE/MSW)

-   So, we calculate F as the ratio of MSG/MSE

-   If F is larger than critical value, we reject the null hypothesis

-   If F is 1 $\rightarrow$ no difference between groups

-   We first simulate three different groups

```{r}
# lets make it women, men and children with their different height
set.seed(43)
n <- 120
# now we build men, women and children. with men normally distributed with average of 175, women average 165 and children average 130

men <- rnorm(n/3, mean=175, sd=10)
women <- rnorm(n/3, mean=165, sd=10)
children <- rnorm(n/3, mean=130, sd=10)
# generate a vector of "men", "women" and "children" with number of time as the number of samples
type <- rep(c("men", "women", "children"), each = n/3)
# combine height of all three groups
height <- c(men, women, children)
# Create data frame
df_h <- data.frame(type=type, height=height)

```

Let's plot it

```{r}
ggplot(df_h, aes(x=type, y= height, color= type)) +
  geom_boxplot() +
  geom_jitter(width = 0.2, height = 0, alpha = 0.5) +
  theme_minimal()
```

## Now we do ANOVA

```{r}
model <- aov(height ~ type, data = df_h)
summary(model)
```

## Post Hoc Analysis

-   So, we now there's a difference, but not sure where?

-   We can use post hoc analysis to compare all groups to each other

-   TukeyHSD (Honestly Significant Difference) is a common post hoc test after ANOVA, which compares all groups

```{r}
TukeyHSD(model)

```

-   When we have a specific group comparison we can use Bonferroni correction

```{r}
pairwise.t.test(df_h$height, df_h$type, p.adjust.method = "bonferroni")
```

# Why do we correct for multiple comparisons?

```{r}
set.seed(42)
# Number of groups and participants
num_groups <- 5
participants_per_group <- 20
simulate_false_positives <- function(){
  data <- data.frame(
    Group = factor(rep(1:num_groups, each = participants_per_group)),
    Value = rnorm(num_groups * participants_per_group, mean = 10, sd = 2)
  )
  
  pairwise_result <- pairwise.t.test(data$Value, data$Group, p.adjust.method = "none")
  false_positive_count <- sum(pairwise_result$p.value < 0.05, na.rm = TRUE)
  return(false_positive_count)
}

# Run simulation 5000 times
simulations <- replicate(5000, simulate_false_positives())

# Average number of false positives
mean(simulations)
hist(simulations, 
     main = "Distribution of False Positives (no correction)", 
     xlab = "Number of False Positives per Simulation")

```

## ANOVA with repeated measures

-   Here we have dependent observations

-   So here the "group" might be the same person at different times

-   In $R$ we use the `aov` function with the `Error` function to specify the repeated measures

-   Let's simulate 20 subjects who were measured before treatment, right after treatment and 1 month after treatment

```{r}
set.seed(42)
n_subjects <- 20

# Simulate data
subject <- factor(1:n_subjects)

# Before treatment (baseline)
before <- rnorm(n_subjects, mean = 50, sd = 10)

# Right after treatment (improvement by around 10 points)
after_immediate <- before - rnorm(n_subjects, mean = 10, sd = 5)

# 1 month after (partial return towards baseline, improvement around 5 points)
after_month <- before - rnorm(n_subjects, mean = 5, sd = 5)

# Combine into data frame
df <- data.frame(
  Subject = subject,
  Before = before,
  After_Immediate = after_immediate,
  After_1Month = after_month
)
```

Plot

```{r}
# boxplot of before, after and 1 month
df_long <- df %>%
  pivot_longer(cols = c(Before, After_Immediate, After_1Month), names_to = "Time", values_to = "Value") %>% 
  mutate(Time = factor(Time, levels = c("Before", "After_Immediate", "After_1Month")))

ggplot(df_long, aes(x = Time, y = Value, color = Time)) +
  geom_boxplot() +
  geom_jitter(width = 0.2, height = 0, alpha = 0.5) +
  theme_minimal()
```

-   Now we can do repeated measures

```{r}
model_repeated <- aov(Value ~ Time + Error(Subject/Time), data = df_long)
summary(model_repeated)
```

We can do posthoc analysis as well

```{r}
pairwise.t.test(df_long$Value, df_long$Time, p.adjust.method = "bonferroni")
```

# That's it for today
